Size matters: Divergent morphology in elliptical bifaces from sites in the American Southeast articulates with distinct local reduction practices

### 2022-08-07 supp/index.Rmd
<edit index>

Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood lithic artefact from the ancestral Caddo area. Elliptical bifaces from Jowell Farm were manufactured using Edwards chert, and are thought to have been acquired by the Caddo through trade or exchange with central Texas groups. Bifaces from Millsap Cache were reportedly manufactured from Kay County flint, providing for an empirical test of whether biface morphology differs as a function of raw material. The sample of elliptical bifaces from each site was also divided into two size classes; one reflective of initial production (large), and the other associated with retouch (small). Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shape in the small class. They were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices. Results demonstrate that elliptical biface shape does not differ by raw material, but size does. This finding suggests a shared mental template, coupled with a size differential. The subsequent analysis of biface morphology by size class demonstrated that size does not differ by raw material in the large class, but does in the small class. This finding supports the argument that elliptical biface morphology diverges through local retouch practices. As expected, greater shape diversity occurs in the small class, where Jowell Farm bifaces were found to be more morphologically diverse than those from Millsap Cache. Distinct local reduction practices are advanced as the driver of extant morphological differences found in elliptical bifaces from the ancestral Caddo area of the American Southeast.

@@

  -   Bifaces _do not differ_ in shape, but **differ significantly** in size by raw material/biogeography.
  -   Allometry (shape change with size) was not significant.
  -   Bifaces in the _large size class_ recovered from Millsap Cache and Jowell Farm do not differ in shape or size, suggesting consistent morphology in bifaces from the _large size class_, supporting arguments for production-based standardization.
  -   Bifaces in the _small size class_ recovered from Millsap Cache and Jowell Farm **differ significantly** in size. Thus, while shape remains consistent, size differs, suggesting---at minimum---locally divergent approaches to retouch.
  -   Morphological disparity in the shape of the _small size class_ is **significantly greater** at Jowell Farm, providing evidence for a more standardized approach to local retouch at Millsap Cache. This finding also provides support for the hypothesis that greater shape diversity is associated with retouch efforts, rather than production.

@@

Components of this analytical work flow were developed and funded by a Preservation Technology and Training grant (P14AP00138) to RZS from the National Center for Preservation Technology and Training (NCPTT), and additional grants to RZS from the Caddo Nation of Oklahoma, National Forests and Grasslands in Texas (15-PA-11081300-033) and the United States Forest Service (20-PA-11081300-074). Funding to analyse the bifaces from the Millsap Cache and Jowell Farm was provided by the Heritage Research Center at Stephen F. Austin State University.

>>>>

Elliptical bifaces from Millsap Cache and Jowell Farm (41AN13) comprise two of the largest samples of this ill-understood lithic artefact from the ancestral Caddo area. Elliptical bifaces from Jowell Farm were manufactured using Edwards chert, and are thought to have been acquired by the Caddo through trade or exchange with central Texas groups. Bifaces from Millsap Cache were reportedly manufactured from Kay County flint, providing for an empirical test of whether biface morphology differs as a function of raw material. The sample of elliptical bifaces from each site was also divided into two size classes; one reflective of initial production (large), and the other associated with retouch (small). Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shape in the small class. They were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices. Results demonstrate that elliptical biface shape does not differ by raw material, but size does, suggesting a shared mental template is maintained regardless of biface size. The subsequent analysis of biface morphology by size class demonstrated that size does not differ by raw material in the large class, but does in the small class. This finding supports the argument that elliptical biface morphology diverges through local retouch practices. As expected, greater shape diversity occurs in the small class, where Jowell Farm bifaces were found to be more morphologically diverse than those from Millsap Cache. Distinct local reduction practices are advanced as the driver of extant morphological differences found in elliptical bifaces from the ancestral Caddo area of the American Southeast.

@@

  -   Bifaces _do not differ_ in shape, but **differ significantly** in size by raw material/biogeography.
  -   Allometry (shape change with size) was not significant.
  -   Bifaces in the _large class_ recovered from Millsap Cache and Jowell Farm do not differ in shape or size, suggesting consistent morphology in bifaces from the _large class_, supporting arguments for production-based standardization.
  -   Bifaces in the _small class_ recovered from Millsap Cache and Jowell Farm **differ significantly** in size. Thus, while shape remains consistent, size differs, suggesting---at minimum---locally divergent approaches to retouch.
  -   Morphological disparity in the shape of the _small class_ is **significantly greater** at Jowell Farm, providing evidence for a more standardized approach to local retouch at Millsap Cache. This finding also provides support for the hypothesis that greater shape diversity is associated with retouch efforts, rather than production.

@@

Components of this analytical work flow were developed and funded by a Preservation Technology and Training grant (P14AP00138) to RZS from the National Center for Preservation Technology and Training (NCPTT), and additional grants to RZS from the Caddo Nation of Oklahoma, National Forests and Grasslands in Texas (15-PA-11081300-033) and the United States Forest Service (20-PA-11081300-074). Funding to analyse the bifaces from Millsap Cache and Jowell Farm was provided by the Heritage Research Center at Stephen F. Austin State University.
